---
title: "RAND Plot Styles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RAND Plot Styles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 120,
  dev = "svg"
)
library(ggplot2)
library(randplot)
```

## Color Palettes

randplot ships three categories of palette, matching RAND's brand guidelines. Use `show_rand_pal()` to preview any palette by name.

### Categorical

Use `RandCatPal` to distinguish discrete, unordered groups. Do not mix colors from multiple palettes in a single chart.

```{r cat-preview}
show_rand_pal("RandCatPal")
```

```{r cat-example}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = RandCatPal) +
  labs(
    title = "Iris sepal dimensions by species",
    x = "Sepal length (cm)",
    y = "Sepal width (cm)"
  ) +
  theme_rand()
```

### Sequential

Sequential palettes are for ordered data (low to high). Lighter colors represent lower values; darker colors represent higher values. Pass the palette to `scale_color_gradientn()` or `scale_fill_gradientn()` for continuous data, or use selected stops with `scale_*_manual()` for discrete ordered groups.

```{r seq-previews}
show_rand_pal("RandSeqBluePal")
show_rand_pal("RandSeqYellowBluePal")
show_rand_pal("RandSeqYellowPurplePal")
show_rand_pal("RandSeqPinkPurplePal")
```

```{r seq-example}
ggplot(mtcars, aes(x = wt, y = mpg, color = hp)) +
  geom_point(size = 3) +
  scale_color_gradientn(colors = RandSeqBluePal) +
  labs(
    title = "Fuel efficiency by weight and horsepower",
    x = "Weight (1,000 lbs)",
    y = "Miles per gallon",
    color = "Horsepower"
  ) +
  theme_rand()
```

For a discrete ordered variable, index into the palette to pick evenly-spaced stops:

```{r seq-discrete-example}
mtcars$cyl_f <- factor(mtcars$cyl)

ggplot(mtcars, aes(x = wt, y = mpg, color = cyl_f)) +
  geom_point(size = 3) +
  scale_color_manual(values = RandSeqBluePal[c(3, 6, 9)]) +
  labs(
    title = "Fuel efficiency by weight and cylinder count",
    x = "Weight (1,000 lbs)",
    y = "Miles per gallon",
    color = "Cylinders"
  ) +
  theme_rand()
```

### Diverging

Diverging palettes are for data centered on a meaningful midpoint (zero, an average, a baseline). The two ends are distinct hues; the center is neutral.

```{r div-previews}
show_rand_pal("RandDivTealOrangePal")
show_rand_pal("RandDivBlueOrangePal")
show_rand_pal("RandDivSteelSalmonPal")
show_rand_pal("RandDivInkSalmonPal")
```

```{r div-example}
set.seed(42)
regions <- c(
  "Northeast", "Midwest", "South", "West",
  "Mountain", "Pacific", "Great Plains", "Southwest"
)
div_data <- data.frame(
  region = factor(regions, levels = regions[order(rnorm(8))]),
  change = c(3.2, -1.8, 5.4, -4.1, 2.0, -0.5, 1.3, -3.7)
)

ggplot(div_data, aes(x = change, y = region, fill = change)) +
  geom_col() +
  geom_vline(xintercept = 0, color = "#333333", linewidth = 0.4) +
  scale_fill_gradientn(
    colors = RandDivTealOrangePal,
    limits = c(-6, 6),
    guide = "none"
  ) +
  labs(
    title = "Change from national average (%)",
    x = NULL,
    y = NULL
  ) +
  theme_rand()
```

---

## OKLCH interpolation with `rand_pal_n()`

The named palettes contain 9 pre-computed stops from RAND's brand guidelines. `rand_pal_n(palette, n)` generates any number of colors by interpolating those stops in OKLCH space — the same perceptually uniform color space the brand team used. Hue is interpolated along the shortest arc, matching the CSS `in oklch shorter hue` behavior.

Use it anywhere you need a different step count — discrete ordinal scales, continuous gradients, or small multiples.

```{r oklch-compare, fig.height = 2}
# 5 discrete stops from a 9-stop sequential palette
show_rand_pal <- randplot::show_rand_pal  # already loaded
par(mfrow = c(3, 1), mar = c(2, 1, 1, 1))

pal5 <- rand_pal_n(RandSeqYellowBluePal, 5)
graphics::image(1:5, 1, as.matrix(1:5), col = pal5,
  xlab = "rand_pal_n(RandSeqYellowBluePal, 5)", ylab = "",
  xaxt = "n", yaxt = "n", bty = "n")

pal3 <- rand_pal_n(RandDivTealOrangePal, 3)
graphics::image(1:3, 1, as.matrix(1:3), col = pal3,
  xlab = "rand_pal_n(RandDivTealOrangePal, 3)", ylab = "",
  xaxt = "n", yaxt = "n", bty = "n")

pal35 <- rand_pal_n(RandDivInkSalmonPal, 35)
graphics::image(1:35, 1, as.matrix(1:35), col = pal35,
  xlab = "rand_pal_n(RandDivInkSalmonPal, 35)", ylab = "",
  xaxt = "n", yaxt = "n", bty = "n")

```

```{r oklch-continuous}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_gradientn(colors = rand_pal_n(RandSeqYellowBluePal, 256)) +
  labs(
    title = "Old Faithful eruption density",
    x = "Waiting time (min)",
    y = "Eruption length (min)"
  ) +
  theme_rand()
```

```{r oklch-diverging-smooth}
ggplot(div_data, aes(x = change, y = region, fill = change)) +
  geom_col() +
  geom_vline(xintercept = 0, color = "#333333", linewidth = 0.4) +
  scale_fill_gradientn(
    colors = rand_pal_n(RandDivTealOrangePal, 256),
    limits = c(-6, 6),
    guide = "none"
  ) +
  labs(
    title = "Change from national average (%) — OKLCH gradient",
    x = NULL,
    y = NULL
  ) +
  theme_rand()
```

---

## Theme

`theme_rand()` applies RAND's visual style to any ggplot. It automatically uses ABC Monument Grotesk if the font has been configured with `rand_font_setup()`, falling back to the system sans-serif otherwise.

```{r theme-example}
monthly <- data.frame(
  month = factor(month.abb, levels = month.abb),
  value = c(12, 15, 22, 28, 35, 41, 44, 43, 36, 27, 18, 13)
)

ggplot(monthly, aes(x = month, y = value)) +
  geom_col(fill = RandCatPal[1]) +
  labs(
    title = "Monthly values",
    subtitle = "A demonstration of theme_rand()",
    caption = "Source: Example data",
    x = NULL,
    y = "Value"
  ) +
  theme_rand()
```
